---
title: "Reticulate Environments & Reproducibility: Conda / Virtualenv for trainDist"
author: "Wayne S. Stewart"
date: "`r format(Sys.Date())`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Reticulate Environments & Reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why environments matter

When your R code calls Python via **reticulate**, you want a **stable** Python interpreter
and a **repeatable** set of Python packages. This vignette shows how to:

- Select a specific interpreter (system Python, `virtualenv`, or `conda`).
- Create and manage environments from R.
- Make your choice **reproducible** for collaborators and CI.
- Troubleshoot common issues.

---

## Inspect the current configuration

```{r}
reticulate::py_config()         # active Python & numpy compatibility (if present)
reticulate::py_discover_config()# scan for discovered Python installations
```

---

## Choosing an interpreter for this project

You can pin the interpreter in three main ways (pick one):

### 1) System Python (quickest for stdlib-only code)

```{r}
Sys.setenv(RETICULATE_PYTHON = "/usr/bin/python3")  # adjust per OS
# Windows example:
# Sys.setenv(RETICULATE_PYTHON = "C:/Python311/python.exe")
```

### 2) A `virtualenv` (works anywhere Python is available)

```{r}
# Create once per machine (if it doesn't exist yet)
reticulate::virtualenv_create("~/.virtualenvs/train")

# Activate for this session (do this *before* importing any Python modules)
reticulate::use_virtualenv("~/.virtualenvs/train", required = TRUE)

# Install Python deps (none needed for trainDist, example only)
# reticulate::py_install("numpy", envname = "~/.virtualenvs/train")
```

### 3) A `conda` environment (great when distributing binaries)

```{r}
# Create once per machine
reticulate::conda_create("r-train", python_version = "3.11")

# Activate for this session (before importing Python)
reticulate::use_condaenv("r-train", required = TRUE)

# Install packages into that env if needed
# reticulate::py_install("numpy", envname = "r-train")
```

> **Tip:** Call `use_virtualenv()` or `use_condaenv()` at the **top of your script** or
> inside `.Rprofile` for the project so that it runs before any Python import occurs.

---

## Project-level reproducibility

### Pin the interpreter for collaborators

Create a project `.Renviron` file with a single line:

```
RETICULATE_PYTHON=/absolute/path/to/python
```

Restart R so the environment variable takes effect.

### Capture R packages with `{renv}` (optional but recommended)

```{r}
# install.packages("renv")
renv::init()         # creates renv.lock with R package versions
renv::snapshot()     # record reticulate/knitr/rmarkdown versions
```

### Continuous Integration (CI) sketch

On GitHub Actions (example steps):

1. Set up R and Pandoc (use r-lib/actions).
2. Install system Python or Miniconda.
3. Create the chosen env and set `RETICULATE_PYTHON`.
4. `R CMD check` with `--as-cran`.

This ensures vignettes that import Python code build on CI.

---

## Using `trainDist` with an environment

```{r}
# Example: use a conda env, then load and call
# reticulate::use_condaenv("r-train", required = TRUE)
library(trainDist)
ptrain(0)
qtrain(c(0.25, 0.5, 0.75))
```

For RNG reproducibility across R and Python:

```{r}
set.seed(1)          # R RNG
x <- rtrain(5, seed = 42)  # Python RNG inside
x
```

---

## Common pitfalls and fixes

- **"ModuleNotFoundError: No module named 'train'"**  
  Load the package first so that `.onLoad` adds `inst/python` to `sys.path`.

- **Interpreter mismatch after changing `RETICULATE_PYTHON`**  
  Restart R; once reticulate initializes a Python session, the choice is sticky.

- **Windows path quirks**  
  Prefer absolute paths, e.g., `C:/Users/<you>/AppData/Local/Programs/Python/Python311/python.exe`.

- **Conda not found**  
  Install [Miniconda](https://docs.conda.io/en/latest/miniconda.html) or use `reticulate::install_miniconda()`.

- **Building vignettes fails on CI**  
  Ensure Python is available on PATH *and* the env is activated before the build step.

---

## Appendix: Quick reference

```{r}
# Inspect
reticulate::py_config(); reticulate::py_discover_config()

# Select
# Sys.setenv(RETICULATE_PYTHON = "/path/to/python")
# reticulate::use_virtualenv("~/.virtualenvs/train", required = TRUE)
# reticulate::use_condaenv("r-train", required = TRUE)

# Manage
# reticulate::virtualenv_create("~/.virtualenvs/train")
# reticulate::conda_create("r-train", python_version = "3.11")
# reticulate::py_install(c("numpy"), envname = "r-train")
```
