---
title: "Calling Python from R with reticulate: The Train Distribution"
author: "Wayne S. Stewart"
date: "`r format(Sys.Date())`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Calling Python from R with reticulate: The Train Distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Overview

This vignette demonstrates how to expose **Python** implementations of probability
functions to **R** using the **reticulate** package. We use a custom *train delay*
distribution with probability density function (PDF)

$$
 f(y) = c\,(25 - y^2),\quad -5 < y < 5,\qquad c = 3/500.
$$

From this we build the familiar **dpqr** quartet:

- `dtrain(x)`: density
- `ptrain(q)`: cumulative distribution function
- `qtrain(p)`: quantile function (inverse CDF)
- `rtrain(n)`: random generation via inverse transform

All four are implemented in **Python** and called from **R**.

---

## Installation

From a shell or an R session in the directory above the package:

```r
# install.packages("devtools")
devtools::install_local("trainDist", upgrade = "never")
```

Load the package:

```{r}
library(trainDist)
```

---

## Quick examples (dpqr)

### Density `dtrain()`

```{r}
# Density at a single point and a vector
dtrain(0)
xs <- seq(-10, 10, length.out = 501)
plot(xs, dtrain(xs), type = "l", lwd = 2, col = 2,
     xlab = "x", ylab = "Density", main = "Train density")
```

### CDF `ptrain()`

```{r}
ptrain(0)
curve(ptrain, from = -10, to = 10, n = 501, lwd = 2,
      xlab = "q", ylab = "F(q)", main = "Train CDF")
```

### Quantile `qtrain()`

```{r}
qtrain(0.75)
qtrain(c(0.25, 0.5, 0.75))
```

### Random generation `rtrain()`

```{r}
set.seed(123)
x <- rtrain(10000)
hist(x, prob = TRUE, breaks = 60, col = "gray", border = NA,
     main = "Random sample vs. density", xlab = "y (minutes late)")
curve(dtrain, add = TRUE, n = 601, lwd = 2, col = 2)
```

---

## How the R \u2192 Python bridge is built

This package uses a lightweight pattern that requires **no external Python packages**:

1. **Place Python code inside the package** at `inst/python/train.py`.
2. On package load, **prepend** that directory to Python's `sys.path` so it can
   be imported. We do this in `R/zzz.R`:

```r
.onLoad <- function(libname, pkgname) {
  if (requireNamespace("reticulate", quietly = TRUE)) {
    py_dir <- system.file("python", package = pkgname)
    reticulate::py_run_string(sprintf(
      "import sys, os; p=r'%s';\nif p not in sys.path: sys.path.insert(0, p)", py_dir))
  }
}
```

3. In your exported R functions, **import and call** the Python functions:

```r
#' @export
dtrain <- function(x) {
  mod <- reticulate::import("train", delay_load = TRUE, convert = TRUE)
  as.numeric(mod$dtrain(x))
}
```

Key points:

- `reticulate::import("train", delay_load = TRUE)` delays import until first use.
- `convert = TRUE` converts Python lists / scalars to R vectors / numerics.
- Keeping Python in `inst/python/` avoids path headaches when users install from source or binary.

---

## Required packages and configuration

### R-side dependencies

- **reticulate** (interface to Python).
- **knitr** and **rmarkdown** (to build this vignette).
- **testthat** (optional, for the included sanity checks).

These are declared in the package `DESCRIPTION` and installed as needed.

### Python-side dependencies

- None required for this package (uses only the Python standard library).

> If you later rely on third-party Python libraries (e.g., NumPy), prefer declaring
> them and managing environments with `reticulate` (see below).

### Choosing a Python interpreter

Use `reticulate::py_config()` to see which Python is being used:

```{r}
reticulate::py_config()
```

Common ways to control the interpreter used by reticulate:

- Set the environment variable `RETICULATE_PYTHON` to an absolute path to the
  desired Python **before** starting R (or before loading reticulate):

```r
Sys.setenv(RETICULATE_PYTHON = "/usr/bin/python3")
# or on Windows: Sys.setenv(RETICULATE_PYTHON = "C:/Python311/python.exe")
```

- Use a **virtualenv** or **conda** environment and point reticulate to it:

```r
reticulate::use_virtualenv("~/.virtualenvs/train", required = TRUE)
# or
reticulate::use_condaenv("r-train", required = TRUE)
```

- You can also create and install Python packages into an environment from R:

```r
# Create if needed, then install (example for numpy if you ever add it)
reticulate::virtualenv_create("~/.virtualenvs/train")
reticulate::py_install("numpy", envname = "~/.virtualenvs/train")
```

> Tip: Call `use_virtualenv()` / `use_condaenv()` *before* importing any Python
> modules to ensure the selection takes effect in the current session.

### Reproducibility and seeding

`rtrain(n, seed = NULL)` forwards an optional seed to Python's RNG. For end-to-end
reproducibility across R and Python, set both seeds if you also use R-side randomness:

```r
set.seed(1)          # R-side
r <- rtrain(5, 42)   # Python-side deterministic given seed
```

---

## Troubleshooting

- **"ModuleNotFoundError: No module named 'train'"**: Ensure the package is loaded
  so that `.onLoad` has run (it inserts `inst/python` into `sys.path`).
- **Wrong Python version**: Run `reticulate::py_config()` and set
  `RETICULATE_PYTHON` (or use `use_virtualenv()`/`use_condaenv()`) before importing.
- **Environment conflicts**: Restart R, clear `.Rprofile` hooks that select other
  environments, and re-run the `use_*env()` calls.
- **Windows builds**: If building vignettes from source, ensure a working `rmarkdown`
  toolchain (pandoc is bundled with RStudio) and that the chosen Python is on PATH.

---

## Appendix: Mathematical details

- Normalizing constant: \(c = 3/500\) so that \(\int_{-5}^{5} c(25 - y^2)\,dy = 1\).
- CDF on \((-5,5)\): \(F(y) = 1/2 + 0.15\,y - y^3/500\); 0 for \(y\le -5\), 1 for \(y\ge 5\).
- Quantiles computed by bisection on \([-5,5]\) to invert \(F\).

---

## Session info

```{r}
sessionInfo()
```
